Community Detection Methods for the Generalized Random Dot Product Graph Model

### 2021-05-17 proposal/proposal.Rmd
uploading slides

# Literature Review

## Generative Graph Models and Community Detection

@@

# Proposed Research

## Connecting Existing Generative Graph Models to the RDPG and GRDPG

As discussed in \S 2.1, all Bernoulli graph models can be expressed as a RDPG or GRDPG model, and in particular, the (associative) SBM is a RDPG with a very specific structure in the latent space. A similar result has been shown for the DCBM \cite{lyzinski2014}. We have also extended this to the PABM. In \S 3.3.1, we show that the PABM is equivalent to the GRDPG model for which the communities correspond to subspaces in the latent configuration. We further show that while the latent configuration is not unique, there exists one in particular for which the subspaces are orthogonal. This leads to two straightforward methods for community detection, one involving computing inner products in the ASE and another which is a slight modification of an existing algorithm for subspace clustering. Finally, we will briefly explore alternative, non-spectral-based methods for community detection in the ASE involving Expectation Maximization, MCMC sampling, and Mean Field Variational Inference.

### Popularity Adjusted Block Model

@@

Since the communities are subspaces in the latent space, we can also use existing methods for subspace clustering on the ASE. Of particular interest is Sparse Subspace Clustering (SSC), which is performed by solving an optimization problem for each observed point in a sample. Given $X \in \mathbb{R}^{n \times d}$ with vectors $x_i^\top \in \mathbb{R}^d$ as rows of $X$, the optimization problem $c_i = \min\limits_{c} \|c\|_1$ subject to $x_i = X_{-i} c$ and $c^{(i)} = 0$ is solved for each $i \in [n]$. The solutions are collected into matrix $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top$ to construct affinity matrix $B = |C| + |C^\top|$. If each $x_i$ lie perfectly on one of $K$ subspaces, $B$ is sparse such that $B_{ij} = 0$ $\forall x_i, x_j$ belonging to different subspaces. Then $B$ can describe a graph with at least $K$ disjoint subgraphs, and if the number of subgraphs is exactly $K$, each subgraph maps onto a subspace.

@@

#### Alternative Approaches to Community Detection

@@

## Proposed Generative Graph Models

@@

### Affine Segments

@@

### Mainfold Learning

@@

Simulations and real data analyses: March 2022

>>>>

# Preliminaries

## Block Models

@@

# The Popularity Adjusted Block Model

As discussed in \S 2.1, all Bernoulli graph models can be expressed as a RDPG or GRDPG model, and in particular, the (associative) SBM is a RDPG with a very specific structure in the latent space. A similar result has been shown for the DCBM \cite{lyzinski2014}. We can also extend this to the PABM. In \S 3.3.1, we show that the PABM is equivalent to the GRDPG model for which the communities correspond to subspaces in the latent configuration. We further show that while the latent configuration is not unique, there exists one in particular for which the subspaces are orthogonal. This leads to two straightforward methods for community detection, one involving computing inner products in the ASE and another which is a slight modification of an existing algorithm for subspace clustering. Finally, we will briefly explore alternative, non-spectral-based methods for community detection in the ASE involving Expectation Maximization, MCMC sampling, and Mean Field Variational Inference.

## Connecting the PABM to the GRDPG

@@

Since the communities are subspaces in the latent space, we can also use existing methods for subspace clustering on the ASE. Of particular interest is Sparse Subspace Clustering (SSC), which is performed by solving an optimization problem for each observed point in a sample. Given $X \in \mathbb{R}^{n \times d}$ with vectors $x_i^\top \in \mathbb{R}^d$ as rows of $X$, the optimization problem $c_i = \arg\min\limits_{c} \|c\|_1$ subject to $x_i = X_{-i} c$ and $c^{(i)} = 0$ is solved for each $i \in [n]$. The solutions are collected into matrix $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top$ to construct affinity matrix $B = |C| + |C^\top|$. If each $x_i$ lie perfectly on one of $K$ subspaces, $B$ is sparse such that $B_{ij} = 0$ $\forall x_i, x_j$ belonging to different subspaces. Then $B$ can describe a graph with at least $K$ disjoint subgraphs, and if the number of subgraphs is exactly $K$, each subgraph maps onto a subspace.

@@

## Alternative Approaches to Community Detection

@@

# Generalizations

@@

## Affine Segments

@@

## Manifolds

@@

Simulations, real data analyses, R package: March 2022

### 2021-05-17 proposal/proposal.tex
uploading slides

\hypertarget{literature-review}{%
\section{Literature Review}\label{literature-review}}

\hypertarget{generative-graph-models-and-community-detection}{%
\subsection{Generative Graph Models and Community
Detection}\label{generative-graph-models-and-community-detection}}

@@

\hypertarget{proposed-research}{%
\section{Proposed Research}\label{proposed-research}}

\hypertarget{connecting-existing-generative-graph-models-to-the-rdpg-and-grdpg}{%
\subsection{Connecting Existing Generative Graph Models to the RDPG and
GRDPG}\label{connecting-existing-generative-graph-models-to-the-rdpg-and-grdpg}}

As discussed in \S 2.1, all Bernoulli graph models can be expressed as a
RDPG or GRDPG model, and in particular, the (associative) SBM is a RDPG
with a very specific structure in the latent space. A similar result has
been shown for the DCBM \cite{lyzinski2014}. We have also extended this
to the PABM. In \S 3.3.1, we show that the PABM is equivalent to the
GRDPG model for which the communities correspond to subspaces in the
latent configuration. We further show that while the latent
configuration is not unique, there exists one in particular for which
the subspaces are orthogonal. This leads to two straightforward methods
for community detection, one involving computing inner products in the
ASE and another which is a slight modification of an existing algorithm
for subspace clustering. Finally, we will briefly explore alternative,
non-spectral-based methods for community detection in the ASE involving
Expectation Maximization, MCMC sampling, and Mean Field Variational
Inference.

\hypertarget{popularity-adjusted-block-model}{%
\subsubsection{Popularity Adjusted Block
Model}\label{popularity-adjusted-block-model}}

@@

Since the communities are subspaces in the latent space, we can also use
existing methods for subspace clustering on the ASE. Of particular
interest is Sparse Subspace Clustering (SSC), which is performed by
solving an optimization problem for each observed point in a sample.
Given \(X \in \mathbb{R}^{n \times d}\) with vectors
\(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
\(c_i = \min\limits_{c} \|c\|_1\) subject to \(x_i = X_{-i} c\) and
\(c^{(i)} = 0\) is solved for each \(i \in [n]\). The solutions are
collected into matrix
\(C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top\) to
construct affinity matrix \(B = |C| + |C^\top|\). If each \(x_i\) lie
perfectly on one of \(K\) subspaces, \(B\) is sparse such that
\(B_{ij} = 0\) \(\forall x_i, x_j\) belonging to different subspaces.
Then \(B\) can describe a graph with at least \(K\) disjoint subgraphs,
and if the number of subgraphs is exactly \(K\), each subgraph maps onto
a subspace.

@@

\hypertarget{alternative-approaches-to-community-detection}{%
\paragraph{Alternative Approaches to Community
Detection}\label{alternative-approaches-to-community-detection}}

@@

While direct maximization of the likelihood or log-likelihood is
NP-complete, it is possible to apply an Expectation Maximization (EM)
algorithm to this problem.

\hypertarget{proposed-generative-graph-models}{%
\subsection{Proposed Generative Graph
Models}\label{proposed-generative-graph-models}}

@@

\hypertarget{affine-segments}{%
\subsubsection{Affine Segments}\label{affine-segments}}

@@

\hypertarget{mainfold-learning}{%
\subsubsection{Mainfold Learning}\label{mainfold-learning}}

>>>>

\hypertarget{preliminaries}{%
\section{Preliminaries}\label{preliminaries}}

\hypertarget{block-models}{%
\subsection{Block Models}\label{block-models}}

@@

\hypertarget{the-popularity-adjusted-block-model}{%
\section{The Popularity Adjusted Block
Model}\label{the-popularity-adjusted-block-model}}

As discussed in \S 2.1, all Bernoulli graph models can be expressed as a
RDPG or GRDPG model, and in particular, the (associative) SBM is a RDPG
with a very specific structure in the latent space. A similar result has
been shown for the DCBM \cite{lyzinski2014}. We can also extend this to
the PABM. In \S 3.3.1, we show that the PABM is equivalent to the GRDPG
model for which the communities correspond to subspaces in the latent
configuration. We further show that while the latent configuration is
not unique, there exists one in particular for which the subspaces are
orthogonal. This leads to two straightforward methods for community
detection, one involving computing inner products in the ASE and another
which is a slight modification of an existing algorithm for subspace
clustering. Finally, we will briefly explore alternative,
non-spectral-based methods for community detection in the ASE involving
Expectation Maximization, MCMC sampling, and Mean Field Variational
Inference.

\hypertarget{connecting-the-pabm-to-the-grdpg}{%
\subsection{Connecting the PABM to the
GRDPG}\label{connecting-the-pabm-to-the-grdpg}}

@@

Since the communities are subspaces in the latent space, we can also use
existing methods for subspace clustering on the ASE. Of particular
interest is Sparse Subspace Clustering (SSC), which is performed by
solving an optimization problem for each observed point in a sample.
Given \(X \in \mathbb{R}^{n \times d}\) with vectors
\(x_i^\top \in \mathbb{R}^d\) as rows of \(X\), the optimization problem
\(c_i = \arg\min\limits_{c} \|c\|_1\) subject to \(x_i = X_{-i} c\) and
\(c^{(i)} = 0\) is solved for each \(i \in [n]\). The solutions are
collected into matrix
\(C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top\) to
construct affinity matrix \(B = |C| + |C^\top|\). If each \(x_i\) lie
perfectly on one of \(K\) subspaces, \(B\) is sparse such that
\(B_{ij} = 0\) \(\forall x_i, x_j\) belonging to different subspaces.
Then \(B\) can describe a graph with at least \(K\) disjoint subgraphs,
and if the number of subgraphs is exactly \(K\), each subgraph maps onto
a subspace.

@@

\hypertarget{alternative-approaches-to-community-detection}{%
\subsection{Alternative Approaches to Community
Detection}\label{alternative-approaches-to-community-detection}}

@@

While direct maximization of the likelihood or log-likelihood is
NP-complete, it is possible to apply an Expectation Maximization (EM)
algorithm to this problem. To set this up,

\hypertarget{generalizations}{%
\section{Generalizations}\label{generalizations}}

@@

\hypertarget{affine-segments}{%
\subsection{Affine Segments}\label{affine-segments}}

@@

\hypertarget{manifolds}{%
\subsection{Manifolds}\label{manifolds}}
